---
title: "ch09-tidyr"
author: "Andrew Enfield"
date: "3/24/2017"
output: pdf_document
---

```{r}
library(tidyverse)
```

```{r}
table1
```

```{r}
table1 %>% 
  mutate(rate = cases / population * 10000)
```
```{r}
# cases/year
table1 %>%
  count(year, wt = cases)
```

# Gathering

```{r}
table4a
```

```{r}
(tidy4a <- table4a %>%
  gather(`1999`, `2000`, key = "year", value = "cases"))
```

```{r}
table4b
```

```{r}
(tidy4b <- table4b %>%
  gather(`1999`, `2000`, key = "year", value = "population"))
```

```{r}
left_join(tidy4a, tidy4b)
```

# Spreading

```{r}
table2
```

```{r}
table2 %>%
  spread(key = type, value = count)
```

# Gather and spread exercises

1. Why are gather and spread not perfectly symmetrical?

```{r}
(stocks <- tibble(
  year = c(2015, 2015, 2016, 2016),
  half = c(1,2,1,2),
  return = c(1.88, 0.59, 0.92, 0.17)
))
```

```{r}
stocks %>% 
  spread(key = year, value = return)
```

```{r}
stocks %>% 
  spread(key = year, value = return) %>%
  gather(`2015`, `2016`, key = year, value = return)

```

```{r}
stocks %>% 
  spread(key = year, value = return) %>%
  gather(`2015`, `2016`, key = year, value = return, convert = TRUE)
```

2. WHy does this fail?

```{r}
table4a %>%
  gather(1999, 2000, key = year, value = cases)
```

```{r}
table4a %>%
  gather(`1999`, `2000`, key = year, value = cases)
```

3. Why does spreading fail? How could you add a new column to fix the problem?

```{r}
people <- tribble(
   ~name,            ~key,    ~value, 
  #-----------------|--------|------
  "Phillip Woods",   "age",   45,
  "Phillip Woods",   "height",186, 
  "Phillip Woods",   "age",   50,
  "Jessica Cordero", "age",   37,
  "Jessica Cordero", "height",156
)
people
```

```{r}
people %>%
  spread(key = key, value = value)
```
Spread wants to give us a row per name, with columns for age and height. If we have multiple data values for a given name and age, how does it know how to pick one?

Not sure how I could fix this by adding a column - seems like I should get rid of one of the age rows, or do something like create multiple keys - age50, age45, age37?

4. Tidy this - spread or gather? what are the variables?

```{r}
preg <- tribble(
  ~pregnant, ~male, ~female, 
  "yes",     NA,    10, 
  "no",      20,    12
)
preg
```

```{r}
preg %>%
  gather(male, female, key = "gender", value = "count")
```

```{r}
preg %>%
  gather(male, female, key = "gender", value = "count") %>%
  spread(key = pregnant, value = count)
```

So, we've just rearranged the initial preg data. Why did we do this? I think because the initial data wasn't tidy, and now we are. Why? Remember the rules

1. Every variable must have its own column. 
2. Every observations must have its own row.
3. Every value must have its own cell. 

Here each observation was a person. (You could argue that we need to de-aggregate the counts and have a row per person, I suppose.) So we want a row per gender? I'll stop here - i need to think some more on this if i want to, because it seems like the arguments i'm going to make here about why this is tidy could equally apply to the original tibble?